---
title: "Reproducible Research: Peer Assessment 2"
author: Nam D. Nguyen
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    keep_md: true
---

## Synopsis

This report analyzes the NOAA storm data set that includes data on severe weather events from the year 1950 and end in November 2011. The two questions analyzed are: 1) which weather events cause the most harm with respect to population health, and 2) which weather events cause the greatest economic consequences. Regarding harm to population health, the total resulting fatalities and total resulting injuries are aggregated for each event type. Regarding economic consequences, the total estimated damages to property and the total estimated damages to crop are aggregated for each event type.

The findings suggest that tornadoes are the most harmful event type overall and by a wide margin. For economic impact, floods and hurricanes result in the most damage overall. Droughts result in the largest damage to crops but not much property damage. Floods also result in the largest amount of property damage, by a wide margin.

## Data Processing

Load libraries and set code chunk defaults.

```{r setup, message = FALSE}
library(readr)
library(R.utils)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE)
```

First, retrieve the data set, if it does not exist.

```{r get-data}
setwd("./")
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
file <- file.path("data", "storm.csv.bz2")
data <- file.path("data", "storm.csv")

# Create data directory
if(!file.exists("data")) {
  dir.create("data")
}

# Retrieve data file
if(!file.exists(data)) {
  download.file(fileUrl, destfile = file, method = "curl")
  bunzip2(file, data)
  dateDownloaded <- date()
  dateDownloaded
}
```

To aid in quicker exploration and analysis, we create a trimmed data file with the relevant columns for analysis, if it does not exist. Then, we load the trimmed data set and display the first six rows.

To optimize the size of the trimmed data file, columns for property and crop damage estimate magnitudes, i.e. "PROPDMGEXP" and "CROPDMGEXP", are first cleaned and saved as factors, instead of character.

```{r load-data, results = 'markup'}
# load data
data_rds <- file.path("data", "trim.RDS")

if(!file.exists(data_rds)) {
  dat <- read_csv(data, col_types = cols(EVTYPE = col_factor(),
                                         PROPDMGEXP = col_character(),
                                         CROPDMGEXP = col_character()))
  # keep relevant columns
  trim <- dat %>%
    select("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP",
           "CROPDMG", "CROPDMGEXP")
}
```

Taking a first look at the data set.

```{r summary}
if(!file.exists("data")) {
  str(trim, give.attr = FALSE)
  summary(trim)
}
```

Recode the damage estimate magnitude columns by looking at the set of possible recorded values for each.

```{r check-dmgexp}
if(!file.exists(data_rds)) {
  message("Possible values for Property Damage Estimate Magnitude")
  print(unique(dat$PROPDMGEXP))
  message("")
  message("Possible values for Crop Damage Estimate Magnitude")
  print(unique(dat$CROPDMGEXP))
}
```

The documentation for the data set does not explain values, aside from "K", "M", and "B", which correspond to thousands, millions, and billions. These values are used to indicate the magnitude for the property and crop damage estimates columns. "H" appears to indicate hundreds, if the notation is followed.

All other values will be recoded to "1" to indicate "identity" or no magnitude modifier for the damage estimates. The estimate magnitude columns will then be save as factors.


```{r damage-clean}
valid_mag <- c("H", "K", "M", "B")

if(!file.exists(data_rds)) {
  trim <- trim %>%
    mutate(PROPDMGEXP = toupper(PROPDMGEXP),
           CROPDMGEXP = toupper(CROPDMGEXP),
           PROPDMGEXP = ifelse(PROPDMGEXP %in% valid_mag, PROPDMGEXP, 1),
           CROPDMGEXP = ifelse(CROPDMGEXP %in% valid_mag, CROPDMGEXP, 1),
           PROPDMGEXP = as.factor(PROPDMGEXP),
           CROPDMGEXP = as.factor(CROPDMGEXP))
}
```

Next, the damage estimate columns are transformed to apply the magnitude modifier for each observation. The magnitude columns are no longer used, and thus are discarded. This transformation will prepare the damage estimates for statistical analysis.

```{r transform-dmg-est}
if(!file.exists(data_rds)) {
  trim <- trim %>%
    mutate(PROPDMGEXP = case_when(PROPDMGEXP == "H" ~ 100,
                                  PROPDMGEXP == "K" ~ 1000,
                                  PROPDMGEXP == "M" ~ 1000000,
                                  PROPDMGEXP == "B" ~ 1000000000,
                                  TRUE ~ 1),
           CROPDMGEXP = case_when(CROPDMGEXP == "K" ~ 1000,
                                  CROPDMGEXP == "M" ~ 1000000,
                                  CROPDMGEXP == "B" ~ 1000000000,
                                  TRUE ~ 1),
           PROPDMG = PROPDMG * PROPDMGEXP,
           CROPDMG = CROPDMG * CROPDMGEXP) %>%
    select(-PROPDMGEXP, -CROPDMGEXP)
}
```

Finally, the processed data is saved as an RDS file and loaded for analysis. The final data file size was reduced from a 536 MB uncompressed to a 31 MB RDS file.

```{r write-load-trim}
if(!file.exists(data_rds)) {
  write_rds(trim, data_rds)
}

df <- read_rds(data_rds)
str(df)
```

## Results

### Most Harmful Weather Events

```
Across the United States, which types of events (as indicated in the EVTYPE
variable) are most harmful with respect to population health?
```

For the first question, the data will be grouped by event type and the fatalities and injuries for each group will be aggregated.

```{r harm-preprocessing}
evtype <- df %>%
  group_by(EVTYPE) %>%
  summarize(FATALITIES_SUM = sum(FATALITIES),
            INJURIES_SUM = sum(INJURIES))
```

Below are the top 11 event types for the most harmful event types. The first table ranks the event types by the total number of fatalities for an event type. The second table is similar to the first table but ranks by total number of injuries.

```{r fatalities-list}
fatalities <- evtype %>%
  arrange(-FATALITIES_SUM, -INJURIES_SUM) %>%
  head(n = 11)
kable(fatalities, caption = "Top 11 Events by Total Fatalities")
```

```{r injuries-list}
injuries <- evtype %>%
  arrange(-INJURIES_SUM, -FATALITIES_SUM) %>%
  head(n = 11)
kable(injuries, caption = "Top 11 Events by Total Injuries")
```

Tornadoes sits at the top of both lists and are by far the most harmful event type.

The two lists share many of the same events, although the order differs. Looking at the differences between the two lists help show the type of events that are cause more fatalities than injuries and vice versa.

```{r harm-fatalities-diff}
setdiff(fatalities$EVTYPE, injuries$EVTYPE)
```

For rip currents and avalanche, the number of fatalities exceeds its number of injuries, which suggests that death is more likely than injuries for these events. This relationship is not true for high winds, and its exclusion from the top injury-causing-events list may indicate that, while it causes a large share of fatalities, it's not as injurious as other events.

```{r harm-injuries-diff}
setdiff(injuries$EVTYPE, fatalities$EVTYPE)
```

The three items that are on the top injury-causing-events list show several stormy events that result in fewer fatalities when compared to the level of injuries that result.

Finally, to get a better idea of where the events that result in the most fatalities and the ones that cause the most injuries sit in relation to each other, the plot below shows each event's total injuries vs. its total fatalities.

Tornado is excluded from the plot because the order of difference between it and other events makes the magnitude of non-tornado events difficult to discern. Tornadoes, to reiterate, result in the most total fatalities and total injuries, by a large margin.

```{r harm-plot, fig.width = 10}
harm <- fatalities %>%
  bind_rows(injuries) %>%
  distinct(EVTYPE, .keep_all = TRUE) %>%
  filter(EVTYPE != "TORNADO") %>%
  arrange(-FATALITIES_SUM)

ggplot(harm, aes(INJURIES_SUM, FATALITIES_SUM)) +
  geom_point() +
  geom_text_repel(aes(label = EVTYPE)) +
  labs(title = "Total Injuries vs. Total Fatalities by Event Type",
       x = "Injuries",
       y = "Fatalities") +
  theme_minimal()
```

### Most Economically Damaging Weather Events

```
Across the United States, which types of events have the greatest
economic consequences?
```

The analysis for greatest economic consequences is similar to the one done for most harmful to human population. See that section for additional details.

The analysis on economic impact will aggregate the total property damage and crop damage by event type.

```{r damage-preprocessing}
damage <- df %>%
  group_by(EVTYPE) %>%
  summarize(PROPDMG_SUM = sum(PROPDMG),
            CROPDMG_SUM = sum(CROPDMG))
```

Below are the lists ranking the top ten event types that cause the most property damage and the top ten event types that cause the most crop damage.

```{r prop-damage-list}
prop_damage <- damage %>%
  arrange(-PROPDMG_SUM, -CROPDMG_SUM) %>%
  head(n = 10)
kable(prop_damage, caption = "Top 10 Events by Total Property Damage Estimates")
```

```{r crop-damage-list}
crop_damage <- damage %>%
  arrange(-CROPDMG_SUM, -PROPDMG_SUM) %>%
  head(n = 10)
kable(crop_damage, caption = "Top 10 Events by Total Crop Damage Estimates")
```

The event types that appear to be the most economically devastating for both property and events are floods and hurricanes.

The plot below will give a better sense of the top event types along both property damage and crop damage dimensions.

```{r damage-plot, fig.width = 10}
damage_merged <- prop_damage %>%
  bind_rows(crop_damage) %>%
  distinct(EVTYPE, .keep_all = TRUE) %>%
  arrange(-PROPDMG_SUM)

ggplot(damage_merged , aes(PROPDMG_SUM, CROPDMG_SUM)) +
  geom_point() +
  geom_text_repel(aes(label = EVTYPE)) +
  labs(title = "Total Property Damage vs. Total Crop Damage by Event Type",
       x = "Property Damage",
       y = "Crop Damage") +
  theme_minimal()
```

By far, the most economically damaging event for crops are droughts, while the most economically damaging event for property is flood. As shown in the lists above, floods are also the second-most economically damaging events for crops.
